Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-09-12 writeup/pnas-word-order-si.tex
add prose

The variant of the Information Bottleneck that we use has been explored in the machine learning literature by \citet{strouse2017deterministic} and dubbed the ``Deterministic Information Bottleneck'' because, in the setting studied by \citet{strouse2017efficient}, it results in codes that are a deterministic function of the information to be expressed.
We use this version of the Information Bottleneck because (1) it has been proposed in previous literature as a generic formalization of efficiency \citep{ferreri2003least}, and (2) it is not clear what would count as the three variables $Y$, $X$, and $\hat{X}$ in our setting. In our setting we have unordered tree structures $T$ to be conveyed, and utterances $U$ representing them. It is not currently clear what would count as a third variable for the application of the standard Information Bottleneck, although we believe such formulations may be fruitful in the future.

@@

\subsection{Distorted graph-based parsers}

@@

We consider the following head-outward generative model.
Verbs generate verb dependents (\emph{xcomp}) and noun dependents (\emph{obj}), independently.
The overall number $N$ of dependents is $NB(1, p_{branching})$, the number of \emph{obj} dependents is $Binom(p_{obj}, N)$.
Nouns can generate verb dependents (\emph{acl}), of number $NB(1, p_{acl})$.

@@

We computed this for different settings of $p_{branching} \in [0, 0.5]$ and $p_{obj} \in [0,1]$, at $p_{acl} \in \{0, 0.3\}$.\footnote{For $p_{acl} = 0.3$, we only computed values for $p_{branching} \leq 0.4$, due to high computational cost on long sentences resulting from $p_{branching} = 0.5$ and $p_{acl} = 0.3$.}
For these settings, we computed the difference in $I[\mathcal{T}, \mathcal{U}]$ between the two grammars.
Results are shown in Figure~\ref{fig:toy-parseability}.

@@

	\caption{Difference in parseability $I[\mathcal{T}, \mathcal{U}]$ (normalized by sentence length) between the correlating and anti-correlating grammars, for $p_{acl} = 0.0$ (left) and $p_{acl} = 0.3$ (right). Positive values indicate greater values of $I[\mathcal{T}, \mathcal{U}]$ for correlating grammars.}\label{fig:toy-parseability}
\end{figure}

>>>>

The variant of the Information Bottleneck that we use has been explored in the machine learning literature by \citet{strouse2017deterministic} and dubbed the ``Deterministic Information Bottleneck'' because, in the setting studied by \citet{strouse2017efficient}, it results in codes that are a deterministic function of the information to be expressed.
We use this version of the Information Bottleneck because (1) it has been proposed in previous literature as a generic formalization of efficiency \citep{ferreri2003least}, and (2) it is not clear what would count as the three variables $Y$, $X$, and $\hat{X}$ in our setting. In our setting we have unordered tree structures $\mathcal{T}$ to be conveyed, and utterances $\mathcal{U}$ representing them. It is not currently clear what would count as a third variable for the application of the standard Information Bottleneck, although we believe such formulations may be fruitful in the future.

@@

\subsection{Distorted graph-based parsers}
\label{sec:distorted}

@@

When we optimize grammars for efficiency, we find that the optimized grammars exhibit dependency length minimization and the Greenbergian word order correlations. To some extent, this result is surprising, because previous functional explanations for DLM (and the Greenbergian correlations, which are predicted by DLM) have been based on the idea of limitations in working memory, and yet our models do not instantiate any explicit working memory pressures. Nor is it the case that our parsers have implicit locality biases, as demonstrated by the experiments in Section~\ref{sec:distorted} above. Our results therefore suggest that DLM and word order correlations might arise purely because they enable tree structures to be better recovered from trees, and/or they make sequences more predictable.

Here we perform some simulation studies to bolster the argument that DLM and word order correlations can enhance the recoverability of tree structures in a generic sense, without any appeal to memory limitations. To do so, we experiment with toy grammars that can be defined to either (1) exhibit word order correlations or (2) not, and we test whether the grammars of type (1) are more or less parseable than the grammars of type (2). We measure parseability using a CYK PCFG parser, thus removing any potential confounds arising from the neural network parsing model.

Our toy grammar consists of the following head-outward generative model.
Verbs generate verb dependents (\emph{xcomp}) and noun dependents (\emph{obj}), independently.
The overall number $N$ of dependents is $NB(1, p_\text{branching})$, the number of \emph{obj} dependents is $Binom(p_{obj}, N)$.
Nouns can generate verb dependents (\emph{acl}), of number $NB(1, p_\text{acl})$.

@@

We computed this for different settings of $p_\text{branching} \in [0, 0.5]$ and $p_{obj} \in [0,1]$, at $p_\text{acl} \in \{0, 0.3\}$.\footnote{For $p_\text{acl} = 0.3$, we only computed values for $p_\text{branching} \leq 0.4$, due to high computational cost on long sentences resulting from $p_\text{branching} = 0.5$ and $p_\text{acl} = 0.3$.}
For these settings, we computed the difference in $I[\mathcal{T}, \mathcal{U}]$ between the two grammars.

@@

	\caption{Difference in parseability $I[\mathcal{T}, \mathcal{U}]$ (normalized by sentence length) between the correlating and anti-correlating grammars, for $p_\text{acl} = 0.0$ (left) and $p_\text{acl} = 0.3$ (right). Positive values indicate greater values of $I[\mathcal{T}, \mathcal{U}]$ for correlating grammars.}\label{fig:toy-parseability}
\end{figure}

Results are shown in Figure~\ref{fig:toy-parseability}. For all parameter regimes, the correlating grammars have better parseability than the anti-correlating grammars. This is especially the case for grammars with high $p_\text{branching}$.
